---
title: "Florianopolis"
author: "Pablo Almaraz"
output:
  html_document:
    code_folding: show
    df_print: paged
    highlight: tango
    number_sections: no
    theme: united
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---
<style type="text/css">

body, td {
   font-size: 15px;
}
code.r{
  font-size: 12px;
}
pre {
  font-size: 12px
}
</style>

```{r setup, include=FALSE}
if (!require(knitr)) install.packages('knitr')
# knitr::opts_chunk$set(echo = T, message = F, warning = F, cache = F)
knitr::opts_knit$set(root.dir = "./")
library(tidyverse)
library(readxl)
library(mvgam)
library(gratia)
```


```{r Load data, message=FALSE, warning=FALSE, include=FALSE}

source('R/functions.R')

#Load data
# df <- read_excel("data-raw/Dados Historicos Lagoa_Pablo.xlsx")
df <- read_excel("data-raw/Dados Historicos Lagoa_Pablo.xlsx")
# df <- read_delim("data-raw/Dados Historicos Lagoa_Pablo.csv", escape_double = FALSE, trim_ws = TRUE)

# Calcular el promedio mensual por cada región (no importa si faltan días en algunas regiones)
filtered_data <- df %>%
  group_by(Ano, mes) %>%
  summarise_if(is.numeric, mean, na.rm = TRUE) %>%
  ungroup() %>%
  complete(Ano = 2001:2024, mes = 1:12) %>%
  mutate_if(is.numeric, ~replace(., is.nan(.), NA)) %>%
  filter(!(Ano == 2024 & mes > 3))

glimpse(filtered_data)
filtered_data$od = log(filtered_data$`OD%`+1)
filtered_data$Cloro = log(filtered_data$`cloro ug.L`+1)
filtered_data$temp = log(filtered_data$`Temp (oC)`+1)
filtered_data$pH = log(filtered_data$`pH in`+1)

# floRian_data <- filtered_data[, c('Ano','mes','Cloro', 'temp', 'pH', 'od')] %>%
#   dplyr::mutate(time = seq(1,nrow(filtered_data),1)) %>%
#   tidyr::pivot_longer(
#     cols = c(Cloro),
#     names_to = "series",
#     values_to = "Cloro" ) %>%
#   mutate(series = as.factor(series))
#
# model_data_train = model_data[1:(0.9*nrow(model_data)),]
# model_data_test = model_data[(nrow(model_data_train)+1):nrow(model_data),]


filtered_data %>% dplyr::select(Ano, mes, Cloro, temp, pH, od, TRIX) %>%

  # mvgam requires a 'time' variable be present in the data to index
  # the temporal observations. This is especially important when tracking
  # multiple time series. In the Portal data, the 'moon' variable indexes the
  # lunar monthly timestep of the trapping sessions
  dplyr::mutate(time = seq(1,nrow(filtered_data),1)) %>%

  # We can also provide a more informative name for the outcome variable, which
  # is counts of the 'PP' species (Chaetodipus penicillatus) across all control
  # plots
  # dplyr::mutate(count = od) %>%

  # The other requirement for mvgam is a 'series' variable, which needs to be a
  # factor variable to index which time series each row in the data belongs to.
  # Again, this is more useful when you have multiple time series in the data
  dplyr::mutate(series = as.factor('Chlorophyll-a (mg/m3)')) %>%
  # dplyr::mutate(series = as.factor('TRIX')) %>%
  
  # Select the variables of interest to keep in the model_data
  dplyr::select(time, Ano, mes, series, Cloro, temp, pH, od, TRIX) %>% 
  dplyr::rename('Chlorophyll-a (mg/m3)' = Cloro) -> model_data

model_data %>%
  # Create a 'year_fac' factor version of 'year'
  dplyr::mutate(year_fac = factor(Ano),
                month_fac = factor(mes)) -> model_data

model_data_train = model_data[1:(0.9*nrow(model_data)),]
model_data_test = model_data[(nrow(model_data_train)+1):nrow(model_data),]

```


```{r echo=FALSE, message=FALSE, warning=FALSE}
plot_mvgam_series(data = model_data, series = 1, y = 'Chlorophyll-a (mg/m3)')
```

# Model


```{r message=FALSE, warning=FALSE, include=FALSE}

# Rename for sanity...
model_data$Cloro = model_data$`Chlorophyll-a (mg/m3)`

model1 <- mvgam(Cloro ~
                  #s(Ano) +
                  #s(mes,
                  te(mes, Ano,
                  # Cyclic and marginal bases
                  bs = c('cc', 'tp'),
                  # bs = c('cc'),
                  # Reasonable complexities
                  k = c(8, 15)),
                  # Define where the seasonal ends should 'join'
                  knots = list(mes = c(0.5, 12.5)),
                family = gaussian(),
                data = model_data,
                # data = model_data_train,
                # newdata = model_data_test,
                # noncentred = TRUE,
                trend_model = 'GP')

dates <- seq(from = as.Date("2001-01-01"), to = as.Date("2024-03-01"), by = "month")
n_points <- length(dates)  # Total de puntos (279)
# Seleccionar los índices correspondientes a enero de cada año
year_labels <- format(dates, "%Y")  # Extraer los años de las fechas
unique_years <- unique(year_labels)  # Años únicos
indices <- match(unique_years, year_labels)  # Índices correspondientes a enero de cada año

```

# Results

```{r echo=FALSE, message=FALSE, warning=FALSE}

hc <- hindcast(model1, type = 'link')
plot(hc, hide_xlabels=T)

# Añadir el eje x con etiquetas de años
axis(1, at = indices, labels = unique_years)
mtext("Year", side = 1, line = 3)

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

plot_mvgam_fc_green(model1, hide_xlabels=T)

# Añadir el eje x con etiquetas de años
axis(1, at = indices, labels = unique_years)
mtext("Year", side = 1, line = 3)

```

Temporal trend of chlorophyll-a in accorande to the observed hindcast.

```{r echo=FALSE, message=FALSE, warning=FALSE}

plot_mvgam_trend_green(model1, type = 'trend', hide_xlabels=T)

# Añadir el eje x con etiquetas de años
axis(1, at = indices, labels = unique_years)
mtext("Year", side = 1, line = 3)

```

Adjusment of the tensorial model with estimated chlorophyll-a value as a response surface. The increase of chlorophyll-a from 2021 is clear, mainly between May and December.

```{r echo=FALSE, message=FALSE, warning=FALSE}
draw(model1$mgcv_model) + scale_x_continuous(breaks = 1:12,
                       labels = c('J', 'F', 'M', 'A', 'M', 'J', 'J', 'A', 'S', 'O', 'N', 'D')) + 
  labs(x = "Month", y = "Year", title = "Tensor product smooth of month and year") 
```

The conditional effects of chlorophyll-a abundance shows the insights from seasonality and annual trends. This figure shows the annual impact every 5 years, highlighting a change in the chlorophyll-a dynamic from 2021 onwards.

```{r echo=FALSE, message=FALSE, warning=FALSE}

conditional_effects(model1)

```

Residuals and pACF show no significant delays, showcasing the high fitting of the model.
```{r echo=FALSE, message=FALSE, warning=FALSE}

plot(model1, type = 'residuals')

```

The posterior distribution of the estocastic model shows how the model is able to describe the trends in chlorophyll-a in the coastal lagoon.

```{r echo=FALSE, message=FALSE, warning=FALSE}
pp_check(object = model1, ndraws=1000, group = "series")

```

# Conclusions

A change in chlorophyll-a dynamic and be infered from 2021, with an drastic increase of chlorophyll-a concentration and a more dynamic fluctuation until nowadays. This data showcase the impact of the massive wastewater discharge event, turning the ecosystem more unstable and vulnerable. 

```{r eval=FALSE, include=FALSE}
change <- Deriv(model1[["mgcv_model"]], n = nrow(model_data_train))
plot.ts(change$mes$deriv)
plot.ts(change$mes$se.deriv)
m2.dci <- confint(change$gamModel, parm = 'te(mes,Ano)')
#   p2 <- predict(fit_GAM, newdata = single_spat_point[[i]], type = "terms", se.fit = TRUE)
#   m2.dsig <- signifD(p2$fit[,c("s(time)")],
#                      d = m2.d[[Term]]$deriv,
#                      m2.dci[[Term]]$upper,
#                      m2.dci[[Term]]$lower)
#   plot.Deriv(m2.d, sizer = TRUE)

# Cargar librería
library(ggplot2)
df=cbind.data.frame(deriv=change$mes$deriv, se.deriv=change$mes$se.deriv)
# Crear gráfico
ggplot(df, aes(x = 1:nrow(df), y = deriv)) +
  geom_line(color = "blue") + # Línea de la serie temporal de 'deriv'
  geom_ribbon(aes(ymin = deriv - 1.983038*se.deriv, ymax = deriv + 1.983038*se.deriv), fill = "lightblue", alpha = 0.3) + # Área sombreada
  labs(x = "Time", y = "Deriv", title = "Serie Temporal de Deriv con Intervalo de Error") +
  theme_minimal()

```

